\hspace{1cm} Do syntactic and semantic similarity lead to interference effects? \newline Evidence from self-paced reading and event-related potentials using German

### 2023-12-29 paper/main.tex
minor edits

\subsection{Discussion}
We presented the to date largest-sample self-paced reading study that investigated the use of syntactic and semantic features during subject-verb dependency formation. Reading times at the critical verb showed clear evidence for semantic interference, but evidence against syntactic interference. The semantic interference effect was already present at the pre-critical region; i.e., semantic interference occured even before the retrieval event could begin.

@@

At the distractor (which occurred well before the pre-critical region), high semantic interference (animate distractors) lead to longer reading times than low semantic interference (inanimate distractors, see Figure \ref{fig:whole_sentence} A and B). One might argue that this pattern could be due to differences in word length. The animate distractors in our materials were on average 1.1 letters longer than the inanimate distractors (mean length of animate distractors: 9.3, sd: 2.6, mean length of inanimate distractors: 8.2, sd: 2.9). It seems unlikely that this small difference in word length caused such large and long-lasting effects, but in order to be conservative we do not try to interpret this difference.

@@

\textcite{mertzen} observed both syntactic and semantic interference effects in the pre-critical region. 
They discuss several alternative explanations for their observed effects. In addition to encoding interference, \textcite{mertzen} propose three other alternative explanations for their data: parafoveal-on-foveal effects, sentence structure confounds across conditions, and predictive processing effects. Our results cannot be explained by the parafoveal-on-foveal explanation, because there is no parafoveal preview in self-paced reading. The observed difference between the high and low semantic interference conditions is independent of the syntactic manipulation. Therefore, the sentence structure confound is also not a good explanation for the effect in our study. Regarding the predictive-processing explanation, \citet{mertzen} argued that the pre-critical adverb must attach to the upcoming verbal phrase; this leads to an anticipatory creation of a verb phrase chunk in memory, which triggers a retrieval of the subject already at the pre-critical region. However, in our study, the effect started at the distractor and became smaller as the two pre-critical adverbs were read (see Figure \ref{fig:whole_sentence} A and B). The predictive-processing explanation would incorrectly predict that the effect begins at the first adverb, which is the first word of the verbal phrase. Consequently, the prediction explanation is also ruled out in our data, rendering encoding interference the most likely explanation of the effects found prior to the critical verb in our data. Crucially this line of reasoning should not be limited to the explanation of the pre-critical effects, but should be extended to the effects in the critical region as well. The reading times for high vs.\ low semantic interference conditions started to differ at the distractor and differed throughout the rest of the sentence. It is not possible to pinpoint whether the effect in the critical region is due to a retrieval initiated at this region, or still due to encoding and maintaining the distractor in memory \citep{ness2017}, or a combination of encoding and retrieval interference. Because of the unclear status of the effects at the critical verb, we refrain from interpreting the interaction which was only observed in this region.

@@

The contrast between our results and those from  \citeauthor{mertzen}'s (\citeyear{mertzen}) German experiment is surprising given that both used German items and partially even identical items. So, why did we find a semantic interference effect in all regions, whereas \citet{mertzen} only found one in the pre-critical region? One possible explanation is lower power in \citeauthor{mertzen}'s (\citeyear{mertzen}) study. While they already had considerably more participants and items than many psycholinguistic studies, in the present study the number of participants was much higher than in theirs (774 vs.\ 121) and the number of items was also higher (60/120 vs.\ 40). 
%This is also supported by the reduced uncertainty of the individual-level estimates in our data compared to \citeauthor{mertzen}'s (\citeyear{mertzen}). 
We will address the absence of syntactic interference in our data in the general discussion.

@@

\begin{table}[]
    \caption{Relatively informative priors for the analysis of the event-related potentials \citep{nicenboim_stats}. Priors on the slope parameters were truncated to an upper bound of zero, i.e., only negative values. This corresponds to more negative brain responses to high vs.\ low interference conditions in line with the literature \citep{lee_garnsey, martinetal2014, schoknecht2022, vasishth_drenhaus_2011}. The standard deviations of the slope priors were varied in order to conduct a Bayes factor sensitivity analysis. See the corresponding assumed a priori range of the difference between high vs.\ low interference.}
    \label{tab:eeg_priors}
    \centering
    \begin{tabular}{lllr}
    \toprule
    Parameter&Prior & &Assumed Range ($\mu V$)\\
    \midrule
  Intercept & Normal(2, 5)& & [-8, 12]\\
  \cmidrule{2-4}
  \multirow{4}{1cm}{slope} & Normal(0, 0.1) &\hspace{-1em}\rdelim\}{4}{*}[upper bound: 0]& [-0.2, 0]\\
  & Normal(0, 0.5)& & [-1, 0]\\
  & Normal(0, 1) && [-2, 0]\\
  & Normal(0, 5) && [-11, 0]\\
  \cmidrule{2-4}
  sigma & Normal(10, 5)& \hspace{-1em}\rdelim\}{2}{*}[lower bound: 0]\\
  SD & Normal(0, 2)&\\
    \bottomrule
    \end{tabular}
\end{table}

Models were run with four chains and 20,000 iterations of which 2,000 were used as warm-up phase \citep{schad_etal_2022_BF}. For the calculation of Bayes factors and the corresponding sensitivity analysis, we defined a range of priors on the parameters of interest, assuming a range of effect sizes \citep{nicenboim_stats}. We chose slope priors which assume a wide range of effect sizes from $-0.2$ to $-11$\,$\mu V$. However, the more realistic a priori effect sizes probably range from $-0.2$ to at most $-2$\,$\mu V$. This assessment is based on our expectation that the interference effect investigated here, being a subtle manipulation, is probably smaller in magnitude than the well-established  effect of cloze probability on the N400 amplitude analyzed by \textcite{nicenboim_stats}, which ranges from $0$ to $5$\,$\mu V$ \parencite[see section 5.2.5][]{nicenboim_stats}.

>>>>

\subsection{Discussion}
We presented the to-date largest-sample self-paced reading study that investigated the use of syntactic and semantic features during subject-verb dependency formation. Reading times at the critical verb showed clear evidence for semantic interference, but evidence against syntactic interference. The semantic interference effect was already present at the pre-critical region; i.e., semantic interference occurred even before the retrieval event could begin. The effects in the pre-critical region are discussed next.

@@

At the distractor (which occurred well before the pre-critical region), high semantic interference (animate distractors) lead to longer reading times than low semantic interference (inanimate distractors, see Figure \ref{fig:whole_sentence} A and B). One might argue that this pattern could be due to differences in word length. The animate distractors in our materials were on average 1.1 letters longer than the inanimate distractors (mean length of animate distractors: 9.3, sd: 2.6, mean length of inanimate distractors: 8.2, sd: 2.9). It seems unlikely that this small difference of one letter in word length caused such large and long-lasting effects, but of course one cannot rule out this possibility with complete certainty.

@@

\textcite{mertzen} observed both syntactic and semantic interference effects in the pre-critical region. 
They discuss several alternative explanations for their observed effects. In addition to encoding interference, \textcite{mertzen} propose three other alternative explanations for their data: parafoveal-on-foveal effects, sentence structure confounds across conditions, and predictive processing effects. Our results cannot be explained by the parafoveal-on-foveal explanation, because there is no parafoveal preview in self-paced reading.

Regarding the possibility that sentence structure confounds as a possible explanation for the effects in the pre-critical region, this also seems implausible because the observed difference between the high and low semantic interference conditions is independent of the syntactic manipulation; it is the syntactic manipulation that has the confound. Therefore, the sentence structure confound is also not a good explanation for the semantic interference effect observed in our study.

Regarding the predictive-processing explanation, \citet{mertzen} argued that the pre-critical adverb must attach to the upcoming verbal phrase; this leads to an anticipatory creation of a verb phrase chunk in memory, which triggers a retrieval of the subject already at the pre-critical region. However, in our study, the effect started at the distractor and became smaller as the two pre-critical adverbs were read (see Figure \ref{fig:whole_sentence} A and B). The predictive-processing explanation would incorrectly predict that the effect begins at the first adverb, which is the first word of the verbal phrase. Consequently, the prediction explanation is also ruled out in our data, rendering encoding interference the most likely explanation of the effects found prior to the critical verb in our data. Crucially, this line of reasoning should not be limited to the explanation of the pre-critical effects, but should be extended to the effects in the critical region as well. The reading times for high vs.\ low semantic interference conditions started to differ at the distractor and differed throughout the rest of the sentence. It is not possible to pinpoint whether the effect in the critical region is due to a retrieval initiated at this region, or still due to encoding and maintaining the distractor in memory \citep{ness2017}, or a combination of encoding and retrieval interference \parencite{Yadavetal2022}. Because of the unclear status of the effects at the critical verb, we refrain from interpreting the interaction which was only observed in this region.

@@

There is one further possible reason for the diverging effects in our studies compared to previous work: the theoretically-predicted effects we are investigating here may be too subtle to consistently be detectable across studies. In several other attempts to replicate previously observed findings, we have often found it impossible to replicate the originally claimed effects, despite spending four or more years collecting data to obtain sufficient evidence. For example, contrary to published claims \parencite[e.g.,][]{vandyke_mcelree06}, in a large cross-linguistic study, we found evidence against proactive interference from sentence-external distractors \parencite{mertzen_repVDM2006}. A second example is the well-known syntactic local coherence effect in English \parencite{taboretal04}; in a recent large-sample study, we have found evidence against syntactic local coherence \parencite{lcpaape2023}.  A third example is the interaction of locality and surprisal effects reported in \textcite{levy&keller_2013}; we attempted and failed to replicate the key claims in seven experiments \parencite{vasishth2018_signficancefilter}. A fourth example is the processing of antecedent-reflexive dependencies; \textcite{dillon2013} argued that reflexives are exempt from interference, but a large-scale direct replication attempt  \parencite{jaeger_etal_2020} found evidence that reflexives do show interference effects \parencite[cf.][]{yadav2021individual}. Finally, we have also failed to replicate the surprising claim  \parencite{gibsonwu,hsiao03} that Chinese object relatives are easier to process than subject relatives \parencite{VasishthetalPLoSOne2013,VasishthetalPLoSOne2013}. It is in principle possible that all the originally claimed effects actually represent true effects but, due to the inherent variability of data, the effects may be impossible to detect consistently in replications.  Independent of any power issues in all these different studies, it is possible that it is unreasonable to expect consistent replications of effects in behavioral data, especially when investigating subtle effects.  Certain other effects in psycholinguistics, such as strong garden-path effects \parencite{frazierrayner82}, are easily and consistently replicable \parencite[e.g.,][]{huang2023surprisal}. By contrast, subtle manipulations such as interference effects and local coherence may not show up consistently in replication attempts. One strong possibility is that there are systematic and important individual differences between participants' behavior; an increased focus away from investigating average effects and toward individual differences may reveal important insights about the conditions under which subtle effects are replicable; we demonstrate this point for the antecedent-reflexive dependency in \parencite{yadav2021individual}.

Nevertheless, the contrast between our results and those from  \citeauthor{mertzen}'s (\citeyear{mertzen}) German experiment is surprising given that both used German items and partially even identical items. So, why did we find a semantic interference effect in all regions, whereas \citet{mertzen} only found one in the pre-critical region? Apart from the methodological difference (eye-tracking vs.\ self-paced reading), a plausible  explanation for the divergence is lower power in \citeauthor{mertzen}'s (\citeyear{mertzen}) study. While they already had considerably more participants and items than many psycholinguistic studies, in the present study the number of participants was much higher than in theirs (774 vs.\ 121) and the number of items was also higher (60/120 vs.\ 40). 
We will revisit the absence of syntactic interference in our data in the general discussion, where we investigate the question using computational modeling.

We turn next to the event-related potentials experiment.

@@

\begin{table}[!htbp]
    \caption{Relatively informative priors for the analysis of the event-related potentials \citep{nicenboim_stats}. Priors on the slope parameters were truncated to an upper bound of zero, i.e., only negative values. This corresponds to more negative brain responses to high vs.\ low interference conditions in line with the literature \citep{lee_garnsey, martinetal2014, schoknecht2022, vasishth_drenhaus_2011}. The directional priors also serve to allow us to quantify the extent to which a predicted effect replicates; the Bayes factors can be interpreted as indicating the evidence supporting or not supporting replication success \citep{verhagen2014bayesian}.
    The standard deviations of the slope priors were varied in order to conduct a Bayes factor sensitivity analysis. See the corresponding assumed a priori range of the difference between high vs.\ low interference.}
    \label{tab:eeg_priors}
    \centering
    \begin{tabular}{lllr}
    \toprule
    Parameter&Prior & &Assumed Range ($\mu V$)\\
    \midrule
  Intercept & Normal(2, 5)& & [-8, 12]\\
  \cmidrule{2-4}
  \multirow{4}{1cm}{slope} & Normal(0, 0.1) &\hspace{-1em}\rdelim\}{4}{*}[upper bound: 0]& [-0.2, 0]\\
  & Normal(0, 0.5)& & [-1, 0]\\
  & Normal(0, 1) && [-2, 0]\\
  & Normal(0, 5) && [-11, 0]\\
  \cmidrule{2-4}
  sigma & Normal(10, 5)& \hspace{-1em}\rdelim\}{2}{*}[lower bound: 0]\\
  SD & Normal(0, 2)&\\
    \bottomrule
    \end{tabular}
\end{table}

Bayesian models were run with four chains and 20,000 iterations of which 2,000 were used as warm-up phase \citep{schad_etal_2022_BF}. For the calculation of Bayes factors and the corresponding sensitivity analysis, we defined a range of priors on the parameters of interest, assuming a range of effect sizes \citep{nicenboim_stats,schad_etal_2022_BF}. We chose slope priors which assume a wide range of effect sizes from $-0.2$ to $-11$\,$\mu V$. However, the more realistic a priori effect sizes probably range from $-0.2$ to at most $-2$\,$\mu V$. This assessment is based on our expectation that the interference effect investigated here, being a subtle manipulation, is probably smaller in magnitude than the well-established  effect of cloze probability on the N400 amplitude analyzed by \textcite{nicenboim_stats}, which ranges from $0$ to $5$\,$\mu V$ \parencite[see section 5.2.5][]{nicenboim_stats}.
